### 2018-02-18 doc/manuscript.Rmd
abbreviations and fixes to abstract

Abbreviations:

- TG
- TGFA
- IS
- PROMISE
- ISI
-

@@

**Aims/hypothesis:**
Although elevated serum triacylglyceride (TG) is a well-described risk factor for
type 2 diabetes (T2D), few data are available regarding the role of the specific fatty
acid (FA) composition within serum TG (TGFA) in the pathogenesis of T2DM. Our aim, therefore, was to
examine longitudinal associations of TGFA with insulin sensitivity (IS) and
beta-cell function.

**Methods**:
We used longitudinal data (3 visits over 6 years) from the Prospective
Metabolism and Islet Cell Evaluation (PROMISE) cohort of adults (n=477) who were at-risk
for diabetes at baseline. Glucose and insulin from an OGTT were used to calculate the outcome variables
Matsuda index (ISI), HOMA2-%S, Insulinogenic Index over HOMA-IR (IGI/IR),
and Insulin Secretion-Sensitivity Index-2 (ISSI-2). Gas chromatography
quantified TGFA composition. Generalized estimating equations (GEE) adjusted for
confounders and partial least squares (PLS) were used for the analysis.

**Results**: 
The metabolic outcome variables declined by 14% to 27% over the 6-years. In the adjusted
GEE models, four TGFA (14:0, 16:0, 14:1n-7, 16:1n-7 as mol%) had
strong negative associations with IS while others (e.g. 18:1n-7, 18:1n-9, 20:2n-6, 
20:5n-3) had strong positive associations. Few associations were seen for 
beta-cell function, except for 16:0 (negative) and 18:1n-7 (positive). PLS
analysis indicated four TGFA (14:0, 16:0, 14:1n-7, 16:1n-7) that are markers of de novo lipogenesis (DNL)
clustered together and strongly predicted
lower IS. These four TGFA also correlated highly (r>0.4) with clinically
measured TG.

**Conclusions**:
We found that higher proportions of a cluster of four DNL TGFA 
strongly predicted lower IS as well as hypertriglyceridemia.

>>>>

# Abbreviations:

- TGFA: Triacylglycerol fatty acids
- IS: Insulin sensitivity
- PROMISE: Prospective Metabolism and Islet Cell Evaluation cohort
- ISI: Insulin sensitivity index
- GEE: Generalized estimating equations
- PLS: Partial least squares
- HOMA2-%S: Homeostatic model of assessment 2 - percent sensitivity
- IGI/IR: Insulinogenic index over homeostatic model of assessment for insulin
resistance
- ISSI-2: Insulin Secretion-Sensitivity Index-2
- DNL: De novo lipogenesis
- ALT: Alanine aminotransferase
- NEFA: Non-esterified fatty acids
- OGTT: Oral glucose tolerance test
- CV: Coefficient of variation
- IFG: Impaired fasting glucose
- IGT: Impaired glucose tolerance
- MAQ: Modified activity questionnaire
- WC: Waist circumference
- MET: Metabolic equivalent of task
- SD: Standard deviation

@@

**Aims:**
Our aim was to examine longitudinal associations of triacylglyceride fatty acid
(TGFA) composition with insulin sensitivity (IS) and beta-cell function.

**Methods**:
Adults at-risk for T2D (n=477) had glucose and insulin measured from a glucose
challenge at 3 time points over 6 years. The outcome variables Matsuda index
(ISI), HOMA2-%S, Insulinogenic Index over HOMA-IR (IGI/IR), and Insulin
Secretion-Sensitivity Index-2 (ISSI-2) were computed from the glucose challenge.
Gas chromatography quantified TGFA composition from the baseline. We used
adjusted generalized estimating equations (GEE) models and partial least squares
(PLS) regression for the analysis.

**Results**: 
In adjusted GEE models, four TGFA (14:0, 16:0, 14:1n-7, 16:1n-7 as mol%) had
strong negative associations with IS while others (e.g. 18:1n-7, 18:1n-9,
20:2n-6, 20:5n-3) had strong positive associations. Few associations were seen
for beta-cell function, except for 16:0 and 18:1n-7. PLS analysis indicated four
TGFA (14:0, 16:0, 14:1n-7, 16:1n-7) that clustered together and strongly related
with lower IS. These four TGFA also correlated highly (r>0.4) with clinically
measured TG.

**Conclusions**:
We found that higher proportions of a cluster of four TGFA strongly related with
lower IS as well as hypertriglyceridemia. Previous research suggests these four
FA are also strongly influenced by de novo lipogenesis, particular from
carbohydrates.
